\documentclass{article}
\usepackage{amsmath, amsthm, amsfonts, amssymb}
\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\numberwithin{thm}{section}

\theoremstyle{plain}
\newtheorem{prop}{Proposition}
\numberwithin{prop}{section}

\theoremstyle{definition}
\newtheorem{defn}{Definition}
\numberwithin{defn}{section}

\theoremstyle{remark}
\newtheorem*{rem}{Remark}

\newtheorem*{cor}{Corollary}

\numberwithin{equation}{section}

\newcommand{\op}{\prime}
\begin{document}
\title{Elementary Topology of $\mathbb{R}^n$}
\author{Amey Joshi}
\maketitle
\section{Metric spaces}\label{s1}
\begin{defn}[Metric space]\label{s1d1}
Let $X$ be a non-empty set. A mapping $d: X \times X \mapsto \mathbb{R}$ is
a metric if, for all $x, y, z \in X$:
\begin{enumerate}
\item $d(x, y) \ge 0$ and $d(x, y) = 0$ iff $x = y$.
\item $d(x, y) = d(y, x)$.
\item $d(x, y) + d(y, z) \ge d(x, z)$.
\end{enumerate}
The set $X$ together with the metric $d$ is called a metric space and is
denoted by $(X, d)$.
\end{defn}

\begin{defn}[$k$-cell]\label{s1d2}
If $a_i < b_i$ for $i = 1, \ldots, k$ then a $k$-cell is the set 
\[
\{x \in \mathbb{R}^k\;:\; x = (x_1, \ldots, x_k)\;\text{ and } a_i \le x_i \le
b_i\}
\]
\end{defn}

\begin{defn}[Open ball]\label{s1d3}
Let $(X, d)$ be a metric space. An open ball of radius $r > 0$ centred at 
point $x \in X$ is the set $B_r(x) = \{y \in Y \;:\; d(y, x) < r\}$.
\end{defn}

\begin{defn}[Convex set]\label{s1d4}
A set $E \subseteq \mathbb{R}^k$ is convex if $x, y \in E \Rightarrow \lambda x
+ (1 - \lambda)y \in E$ for all $\lambda \in (0, 1)$.
\end{defn}
\begin{rem}
The operation $\lambda x + (1 - \lambda)y$ is defined only on a vector space.
Therefore, the idea of convexity is applicable only for metric spaces with a
vector space structure.
\end{rem}

\begin{prop}\label{s1p1}
$k$-cell is convex.
\end{prop}
\begin{proof}
Let $a_i < b_i, i = 1, \ldots, k$ be such that E = $\{x \in \mathbb{R}^k \;:\;
x = (x_1, \ldots, x_k), a_i \le x_i \le b_i, i = 1, \ldots, k\}$ is a $k$-cell.
Let $x, y \in E$ and $\lambda \in (0, 1)$. If $z = \lambda x + (1 - \lambda y$
so that $z_i = \lambda x_i + (1 - \lambda)y_i$.

Consider
\begin{eqnarray*}
\lambda a_i &\le \lambda x_i \le& \lambda b_i \\
(1 - \lambda) a_i &\le (1 - \lambda)y_i \le & (1 - \lambda)b_i.
\end{eqnarray*}
Adding them, we get $a_i \le z_i \le b_i$ or that $z \in E$.
\end{proof}

\begin{prop}\label{s1p2}
An open ball is convex.
\end{prop}
\begin{proof}
Let $x, y \in B_r(x_0)$. Then $|x - x_0| < r, |y - y_0| < r$ so that if $z =
\lambda x + (1 - \lambda)y$ then $|z - x_0| = |\lambda x + (1 - \lambda)y -
\lambda x_0 - (1 - \lambda)x_0| \le \lambda|x - x_0| + (1 - \lambda)|y - x_0|
< \lambda r + (1 - \lambda)r = r$.
\end{proof}

The following definitions are for a metric space $(X, d)$ and $E \subseteq X$.
\begin{defn}\label{s1d5}
A neighbourhood of $p$ is the set of points in the ball $B_r(p)$ for some $r>0$.
\end{defn}

\begin{defn}\label{s1d6}
A point $p$ is a limit point of $E$ if every neighbourhood of $p$ has a point
$q \ne p$, such that $q \in E$.
\end{defn}

\begin{defn}\label{s1d7}
If $p$ is not a limit point of $E$ then it called an isolated point of $E$.
\end{defn}

\begin{defn}\label{s1d8}
$E$ is closed if it contains all its limit points.
\end{defn}

\begin{defn}\label{s1d9}
A point $p$ is an interior point of $E$ if there exists $r > 0$ such that 
$B_r(p) \subseteq E$.
\end{defn}

\begin{defn}\label{s1d10}
$E$ is open if every point of $E$ is an interior point.
\end{defn}

\begin{defn}\label{s1d11}
The complement of $E$ is the set $E^c = \{p \in X \;;\; p \notin E\}$.
\end{defn}

\begin{defn}\label{s1d12}
$E$ is perfect if every point of it is a its limit point.
\end{defn}

\begin{defn}\label{s1d13}
$E$ is bounded if there exists $M > 0$ such that for all $p, q \in E$, 
$d(p, q) < M$.
\end{defn}

\begin{defn}\label{s1d14}
$E$ is dense in $X$ is every point of $X$ is either a limit point of $E$ or a
point in $E$.
\end{defn}

\begin{prop}\label{s1p3}
A neighbourhood is an open set.
\end{prop}
\begin{proof}
Consider the neighbourhood $B_r(p)$ of a point $p$ and let $q \in B_r(p)$. Then
$d(p, q) < r$. Let $r^\ast = r - d(p, q)$. Consider a point $s \in 
B_{r^\ast}(q)$. For such a point,
\[
d(p, s) \le d(p, q) + d(q, s) < d(p, q) + r^\ast = r.
\]
\end{proof}

\begin{prop}\label{s1p4}
If $p$ is a limit point of a set $E$ then every neighbourhood of $p$ contains
infinitely many points of $E$.
\end{prop}
\begin{proof}
Let $N$ be a neighbourhood of a limit point $p$ of the set $E$. Since $N$ is a
neighbourhood, so is $B_r(p) \cap N$ for all $r > 0$. Since $p$ is a limit
point, $(B_r(p) \cap N) \cap (E - \{p\}) \ne \emptyset$.

Let, if possible, $N$ be a finite set so that $N \cap (E - \{p\}) = \{q_1, 
\ldots, q_m\}$. Define,
\[
\delta = \min_{1 \le i \le m} d(p, q_i).
\]
As $N$ is a finite set, $\delta > 0$. If we choose $r = \delta/2$, the set 
$(B_{\delta/2}(p) \cap N) \cap (E - \{p\})$ is empty, contradicting our 
hypothesis that $p$ is a limit point.
\end{proof}

An immediate consequence of this proposition is:
\begin{prop}\label{s1p5}
A finite set has no limit points.
\end{prop}
\begin{proof}
Let $E$ be a finite set. Let, if possible, $p \in E$ be its limit point. Then
evert neighbourhood of $p$ has infinitely many points of $E$. Since $E$ is
finite, no such point exists.
\end{proof}

Recall that the proposition $A \Rightarrow B$ is true if $A$ is false or $B$
is true. As a result,
\begin{prop}\label{s1p6}
A non-empty finite set is closed.
\end{prop}

\begin{prop}\label{s1p7}
Any metric space $(X, d)$ viewed as a subset of itself is both open and closed.
\end{prop}
\begin{proof}
For any $r > 0, p \in X$, $B_r(p) = \{x \in X \;:\; d(x, p) < r\}$. By 
definition $B_r(p) \subset X$. Since $p$ is arbitary, this implies that $X$
is open.

A set is closed if it contains all its limit points. The limit points are
members of the ambient space, in this case, $X$ itself. Thus $X$ contains all
its limit points and therefore it is a closed set.
\end{proof}

\begin{prop}\label{s1p8}
Let $E_\alpha$ be an artibrary collection of sets. Then
\[
\left(\cup_\alpha E_\alpha\right)^c = \cap_\alpha E_\alpha^c.
\]
\end{prop}
\begin{proof}
Let
\[
x \in \left(\cup_\alpha E_\alpha\right)^c.
\]
Then $x \notin E_\alpha, \;\forall\; \alpha$, that is $x \in E_\alpha^c, 
\;\forall\alpha$ or that $x \in \cap_\alpha E_\alpha^c$. Therefore,
\[
\left(\cup_\alpha E_\alpha\right)^c \subseteq \cap_\alpha E_\alpha^c.
\]

Not suppose, $y \in \cap_\alpha E_\alpha^c$ then $y \in E_\alpha^c$ or $y
\notin E_\alpha$ for all $\alpha$. Therefore, $y \notin \cup_\alpha E_\alpha$
so that 
\[
y \in \left(\cup_\alpha E_\alpha\right)^c
\]
and hence
\[
\cap_\alpha E_\alpha^c \subseteq \left(\cup_\alpha E_\alpha\right)^c.
\]
\end{proof}

\begin{prop}\label{s1e9}
Let $E$ be open then $E^c$ is closed.
\end{prop}
\begin{proof}
Let $x$ be a limit point of $E^c$. Therefore, every neighbourhood of $x$ has
points in $E^c$. That is, no neighbourhood of $x$ will be entirely in $E$.
That is, $x$ is not an interior point of $E$. Since $E$ is open, all its points
are interior points. Therefore, $x \notin E$, or that $x \in E^c$.

Since $x$ was an arbitrary limit point of $E^c$, this means that every limit
point of $E^c$ lies in $E^c$.
\end{proof}

\begin{prop}\label{s1e10}
Let $E^c$ be closed then $E$ is open.
\end{prop}
\begin{proof}
Let $x$ be a limit point of $E$. If $x$ is an interior point then $x \in E$.
otherwise, every neighbourhood of $x$ has points in $E^c$, making $x$ a limit 
point of $E^c$ as well. But $E^c$ is closed. Therefore, $x \in E^c$. Thus, 
every limit point of $E$ is either an interior point of $E$ or limit point of
$E^c$, making $E$ an open set.
\end{proof}

We can combine the above two propositions into
\begin{thm}\label{s1t1}
A set is open iff its complement is closed.
\end{thm}

Since the complement of the complement of a set is the set itself, this theeorem
is equivalent to
\begin{thm}\label{s1t2}
A set is closed iff its complement is open.
\end{thm}

The next few theorems are important in justifying the definition of a 
topological space.
\begin{thm}\label{s1t3}
An arbitrary union of open sets is open.
\end{thm}
\begin{proof}
Let $\{E_\alpha\}$ be a collection of open sets. Therefore every point in the
collection is an interior point. Therefore, every point in $\cup_\alpha 
E_\alpha$ is an interior point.
\end{proof}

\begin{thm}\label{s1t4}
An arbitrary intersection of closed sets is closed.
\end{thm}
\begin{proof}
Let $\{E_\alpha\}$ be a collection of closed sets. Let $p$ be a limit point of
$F = \cap_\alpha E_\alpha$. Then $B_r(p) \cap (F - \{p\}) \ne \varnothing \;
\forall r > 0$. Therefore, $B_r(p) \cap (E_\alpha - \{p\}) \ne \varnothing \;
\forall r > 0, \;\forall \alpha$, or that $p$ is a limit point of every 
$E_\alpha$. Since $E_\alpha$ are closed, $p \in E_\alpha\;\forall \alpha$ or
that $p \in F$.

Since $p$ was an arbitrary limit point of $F$, every limit point of $F$ belongs
to $F$, making it a closed set.
\end{proof}

An alternate proof is:
\begin{proof}
Let $\{E_\alpha\}$ be a collection of closed sets. Then $\{E_\alpha^c\}$ are
all open so that $\cup_\alpha E_\alpha^c$ is open or
\[
\left(\cap_\alpha E_\alpha\right)^c
\]
is open, or $\cap_\alpha E_\alpha$ is closed.
\end{proof}

\begin{thm}\label{s1t5}
An intersection of a finite collection of open sets is open.
\end{thm}
\begin{proof}
Let $\{E_1, \ldots, E_n\}$ be a finite collection of open sets. Let $x \in 
E_1 \cap \ldots \cap E_n$. Then $x \in E_1, \ldots, x \in E_n$. As all of them
are open sets, there are positive numbers $r_1, \ldots, r_n$ such that 
$B_{r_i} \subset E_i$ for $i = 1, \ldots, n$. Let $r = \min(\{r_1, \ldots, 
r_n\})$. The minimum of a finite set of positive numbers is positive. Therefore,
$B_r(x) \subset E_i$ for all $i = 1, \ldots, n$ or that $B_r(x) \subset E_1
\cap \ldots \cap E_n$. 

Since $x$ was an arbitrary point in $E_1 \cap \ldots \cap E_n$, we showed that
every point in $E_1 \cap \ldots \cap E_n$ is an interior point, making it an
open set.
\end{proof}

\begin{thm}\label{s1t6}
A union of a finite collection of closed sets is closed.
\end{thm}
\begin{proof}
Let $\{E_1, \ldots, E_n\}$ be a finite collection of closed sets. Therefore,
$\{E_1^c, \ldots, E_n^c\}$ is a finite collection of open sets. By the previous
therefore, $E_1^c \cap \ldots E_n^c = (E_1 \cup \ldots \cup E_n)^c$ is open or
$E_1 \cup \ldots E_n$ is closed.
\end{proof}

\begin{defn}\label{s1d15}
Let $(X, d)$ be a metric space and $E \subset X$. Let $E^\op$ be the set of all
limit points of $E$. Then the closure of $E$ is defined as $\bar{E} = E \cup
E^\op$.
\end{defn}

\begin{thm}\label{s1t7}
Let $(X, d)$ be a metric space and $E \subset X$. Then,
\begin{enumerate}
\item $\bar{E}$ is closed.
\item $E = \bar{E}$ iff $E$ is closed.
\item $\bar{E} \subset F$ for every closed set $F \subset X$ such that $E 
\subset F$. 
\end{enumerate}
From points (1) and (3), we see that $\bar{E}$ is the smallest closed subset of
$X$ that contains $E$.
\end{thm}
\begin{proof}
By definition, $\bar{E}$ has all limit points of $E$. Therefore, $\bar{E}$ is 
closed. If $E$ is closed then $E^\op \subset E$ and hence $\bar{E} = E$. 
Conversely, if $\bar{E} = E$ then $E^\op \subset E$, that is, all limit points
of $E$ belong to $E$, making it a closed set.

Let $F \subset X$ be a closed set such that $E \subset F$. Let $p$ be a limit
point of $E$ so that $p \in \bar{E}$. Since $E \subset F$, and $F$ is closed, 
$p \in F$. Therefore, $E^\op \subseteq F$. If $q \in E$ is not a limit point of
$E$ then as as $E \subset F$, $q \in F$. Thus any point in $E \cup E^\op$ is a
member of $F$ so that $\bar{E} \subseteq F$.
\end{proof}

A slightly more detailed proof of item (3) is
\begin{proof}
Let $F \subset X$ be a closed set such that $E \subset F$. Let $p$ be a limit
point of $E$. Then every neighbourhood of $p$ has points in $E$ other than $p$.
Since $E \subset F$, every neighbourhood of $p$ has points in $F$ other than
$p$, making $p$ a limit point of $F$. Since $F$ is closed, $p \in F$. Thus,
$E^\op \subset F$. If $q \in E$ is not a limit point of $E$ then since $E 
\subset F$, $q \in F$. Thus ant point in $E \cup E^\op$ is a member of $F$ so
that $\bar{E} \subseteq F$.
\end{proof}

\begin{thm}\label{s1t8}
Let $E$ be a non-empty set of real numbers which is bounded above. Let $y = 
\sup{E}$. Then $y \in \bar{E}$. If $y \in E$ then $E$ is closed.
\end{thm}
\begin{proof}
If $y = \sup E$ then for all $x \in E$, $x \le y$ and if $y^\op \ge x$ then
$y^\op \ge y$. Consider the set $B_\epsilon(y) = \{x \in \mathbb{R}\;:\; |x - y|
< \epsilon\} = \{x \in \mathbb{R} \;:\; x - \epsilon < y < x + \epsilon\}$. As
$y = \sup E$, numbers in $[x - \epsilon, y) \subseteq E$. This is true for all
$\epsilon$. Therefore, $y$ is a limit point of $E$ so that $y \in \bar{E}$. If
$y \in E$ then $E$ is closed.
\end{proof}

\begin{defn}\label{s1d16}
Let $(X, d)$ be a metric space and $E \subset Y \subset X$. Then $E$ is open 
relative to $Y$ if for each $p \in E$, there exists $r_p > 0$ such that if
$d(p, q) < r_p$ and $q \in Y$ then $q \in E$.
\end{defn}
\begin{rem}
If $E$ is open relative to $Y \subset X$ is for any $p \in E$, we can find 
$\epsilon_p > 0$ such that $B_{\epsilon_p} \cap Y \subset E$.
\end{rem}

\begin{thm}\label{s1t9}
Let $(X, d)$ be a metric space and $Y \subset X$. $E \subset Y$ is open relative
to $Y$ iff $E = Y \cap G$ for some open subset $G$ of $X$.
\end{thm}
\begin{proof}
Let $p \in E$. Since $p$ is also a member of $G$, there exists $\epsilon_p > 0$
such that $B_{\epsilon_p}(p) \subset G$. Choose $\epsilon_p$ so that 
$B_{\epsilon_p} \subset Y$. Then, for a $q \in Y$ such that $q \in 
B_{\epsilon_p}$, $q$ belongs to both $Y$ and $G$ so that it belongs to $E = Y
\cap G$, making $E$ open relative to $Y$.

Now suppose $E$ is open relative to $Y$. Therefore, for a $p \in E$, if we can
find $\epsilon_p > 0$ such that if $d(p, q) < \epsilon_p$ for a point $q \in Y$,
then $q \in E$. We have enclosed a point $p$ of $E$ in an open ball of radius
$\epsilon_p$ such that all $q \in Y$ lying in the ball, also belong to $E$. 
Define
\[
G = \cup_p B_{\epsilon_p}(p).
\]
Then $G$ is open. Since $p \in B_{\epsilon_p}$ this also means that $E \subseteq
G \cap Y$. Now conside a point $x \in G \cap Y$. Since $x \in G$, it belongs to
at least one open ball centred at a point in $E$. Since $x$ also belongs to $Y$,
and $E$ is open relative to $Y$, $x \in E$. Thus $x \in G \cap Y \Rightarrow
x \in E$ so that $G \cap Y \subseteq E$.
\end{proof}

\section{Compact sets}\label{s2}
\begin{defn}\label{s2d1}
An open cover of a set $E$ is a collection of open sets $\{G_\alpha\}$ such that
$E \subseteq \cup_\alpha G_\alpha$.
\end{defn}

\begin{defn}\label{s2d2}
A set is compact if every open cover of it has a finite sub-cover.
\end{defn}

\begin{thm}\label{s2t1}
Let $E \subset Y \subset X$ then $E$ is compact relative to $X$ iff it is 
compact relative to $Y$.
\end{thm}
\begin{proof}
Let $E$ be compact relative to $Y$. Suppose $\{G_\alpha\}$ is a family of
open subsets of $X$ that cover $E$. Then, since $E \subset Y \subset X$, the
family $\{F_\alpha = Y \cap G_\alpha\}$ has open subsets of $Y$ that cover $E$.
Let $F_{\alpha_1}, \ldots, F_{\alpha_n}$ be a finite subcover of this family.
Then $G_{\alpha_1}, \ldots, G_{\alpha_n}$ is the finte subcover of open subsets
of $X$.

Now suppose $E$ is compact relative to $X$. That is, $G_{\alpha_1} \cup \ldots
\cup G_{\alpha_m}$ is a finite open cover of $E$. Then choose $F_{\alpha_i} = 
Y \cap G_{\alpha_i}$. By theorem \ref{s1t9}, $F_{\alpha_i}$ are open relative 
to $Y$. Since $E \subset Y$, $F_{\alpha_1} \cap \ldots \cap
F_{\alpha_m}$ is a finite cover of $E$ such that each $F_{\alpha_i} \subset Y$.
\end{proof}

Thus, unlike openness (or closeness), compactness is not relative to a set's
environment. It is its intrinsic property.

\begin{thm}\label{s2t2}
Compact subsets of metric spaces are closed.
\end{thm}
\begin{proof}
Let $E$ be a compact subset of a metric space $(X, d)$ and $q$ be an arbitrary
point in $E^c = X \setminus E$. For each point $p \in X$, let choose $0 < r_p
< d(p, q)/2$. Then $\cup_p B_{r_p}(p)$ is an open cover of $E$. Since $E$ is
compact, there exists a finite sub-cover, say $B_{r_1}(p_1), \ldots, 
B_{r_n}(p_n)$ of $E$. Each of the sets $B_{r_i}(q)$ are disjoint with $E$. 
Furthermore, $q$ lies in each one of them and therefore in their intersection.
An intersection of a finite number of open sets is open. Thus $q$ has a 
neighbourhood of points in $E^c$ alone. That is, $q$ is an interior point. Since
$q$ was chosen arbitrarily, this means that every point of $E^c$ is an interior
point, making it an open set.
\end{proof}

\begin{thm}\label{s2t3}
Closed subsets of compact sets are compact.
\end{thm}
\begin{proof}
Let $K$ be a compact set and $E \subset K$ be closed. Let $\{G_\alpha\}$ be
an open cover of $E$. Define $F_\alpha = G_\alpha \cup E^c$. Since $E$ is 
closed, $E^c$ is open. Therefore, $F_\alpha$ is open and $\{F_\alpha\}$ is an
open cover of $K$. Since $K$ is compact, there is a finite sub-cover, say $
F_1, \ldots, F_n$. Let a point $x \in E$ belong to $F_i = G_i \cup E^c$ then it
follows that $x \in G_i$. Therefore, $G_1, \ldots, G_n$ is an open cover of $E$,
making it a compact set.
\end{proof}

Theorem \ref{s2t3} is true in any topological space. However, theorem \ref{s2t2}
needs an underlying metric space.

\begin{prop}\label{s2p1}
Let $(X, d)$ be a metric space, $K$ be a compact subset of $X$ and $F$ be a 
closed subset of $X$ then $F \cap K$ is compact.
\end{prop}
\begin{proof}
$K \subset X$ is compact implies that $K$ is closed by virtue of theorem 
\ref{s2t2}. An intersection of two closed sets is closed. Therefore $F \cap K$
is closed. Since $F \cap K \subset K$, by theorem \ref{s2t3}, $F \cap K$ is
also compact.
\end{proof}

\begin{thm}\label{s2t4}
Let $(X, d)$ be a metric space and $\{K_\alpha\}$ be a family of compact subsets
of $X$ such that the intersection of every finite subfamily is not empty. Then
$\cap K_\alpha$ is not empty.
\end{thm}
\begin{proof}
Let, if possible, $\cap_\alpha K_\alpha$ be empty. Without loss of generality,
let $K_1$ be such that it has no element in common with the rest of the family.
As $K_\alpha$ are compact subsets of metric space, they are closed. Therefore,
$K_\alpha^c$ are open. Furthermore, $K_1$ has elements in common with at least
a few of them, that is $\cup_{\alpha \ne 1} K_\alpha^c$ are an open cover of
$K_1$. As $K_1$ is also compact so that there is a finite sub-cover, say 
$K_{\alpha_1}^c, \ldots, K_{\alpha_m}^c$. But this also means that
\[
K_1 \cap K_{\alpha_1} \cap \ldots K_{\alpha_m}
\]
is empty contradicting our hypothesis that the intersection of no finite 
subfamily of $\{K_\alpha\}$ is empty.
\end{proof}
\begin{rem}
If $K_1$ was not compact then the arbitary intersection $\cap K_\alpha$ will
be empty.
\end{rem}

\begin{prop}\label{s2p2}
If $\{K_n\}$ is a nested family of non-empty compact sets, that is $K_1 \supset 
K_2 \supset \ldots$ then $\cap_{n \ge 1} K_n$ is non-empty.
\end{prop}
\begin{proof}
By the nested nature of the sets, if $\{K_{i_1}, \ldots, K_{i_m}\}$ is any
finite sub-family and if $j = \max\{i_1, \ldots, i_m\}$ then $K_{i_1} \cap
\ldots \cap K_{i_m} = K_j \ne \varnothing$. Therefore, by theorem \ref{s2t4},
$\cap_{n \ge 1}K_n \ne \varnothing$.
\end{proof}

\begin{thm}\label{s2t5}
If $E$ is an infinite subset of a compact set $K$ then $E$ has a limit point
in $K$.
\end{thm}
\begin{proof}
A point $p \in K$ is a limit point of $E$ if for all $\epsilon_p > 0$,
$B_{\epsilon_p}(p) \cap (E - \{p\}) \ne \varnothing$. If no point of $K$ is
a limit point of $E$ then there exists $\epsilon_p > 0$ for which 
$B_{\epsilon_p}(p)$ has at most one point of $E$. The collection of open balls
$B_{\epsilon_p}(p)$ for all $p \in K$ is an open cover of $K$. Since $K$ is 
compact, there exists a finite subset of these balls covering $K$. But each one
of them will have at most one point of $E$, which means $K$ has only a finite
number of points of $E$. This contradicts the hypothesis that $E$ is an infinite
subset of $E$.
\end{proof}

\begin{thm}\label{s2t6}
If $\{I_n\}$ is a sequence of closed intervals $[a_n, b_n]$ of real numbers such
that $I_n \supset I_{n+1}$. Then $\cap_{n \ge 1}I_n \ne \varnothing$.
\end{thm}
\begin{proof}
Since the intervals are nested, we have $a_1 \le a_2 \le \ldots \le b_2 \le b_1$
so that the sequence of real numbers $\{a_n\}$ is monotone increasing and 
bounded from above. Let $x$ be its supremum so that $a_n \le x$ for all $n$. 
Since $[a_1, b_1]$ is closed, the limit of $\{a_n\}$ lies in it. In fact, it
lies in all closed intervals $[a_m, b_m]$ for the same reason. Therefore, their
intersection is not empty.
\end{proof}

\begin{rem}
Consider the sequence of intervals $(0, 1/n)$. Here all $a_n$ are zero and hence
$\sup\{a_n\} = 0$ but $0$ is not a member of any interval. Therefore, their
intersection is empty.

On the other hand the intersection of intervals $[0, 1/n)$ is $\{0\}$.
\end{rem}

We will extend the previous theorem to $k$-cells in $\mathbb{R}^k$.
\begin{thm}\label{s2t7}
Let $k$ be a positive integer. If $\{I_n\}$ is a sequence of $k$-cells such
that $I_n \supset I_{n+1}$ got all $n \in \mathbb{N}$ then $\cap_{n \ge 1}I_n
\ne \varnothing$.
\end{thm}
\begin{proof}
The cell $I_n$ is defined as the set of points $x \in \mathbb{R}^k$ such that
$a^n_i \le x_i \le b^n_i$, for $i = 1, \ldots, k$. We thus have $k$-sequences
$\{a^n_i\}$ for $i = 1, \ldots, k$, each one of which is monotone increasing.
Since all of them are bounded from above, their suprema $x^\ast_i$ exist. Since
every $k$ cell is closed, the limit points of these sequences lie in every cell.
As a result, their intersection is non-empty.
\end{proof}

\begin{thm}\label{s2t8}
The interval $[a, b]$ is compact.
\end{thm}
\begin{proof}
Let $\{G_\alpha\}$ be an open cover of $[a, b]$. Define the set
\[
E = \{x \in [a, b]\;:\; [a, x]\text{ is covered by finitely many} G_\alpha.\}.
\]
Since $G_\alpha$ is an open cover of $[a, b]$, there is an open set, say $G_1$
that has has $a$ in it. Thus $a \in E$. We also note that $E$ is bounded. A
non-empty bounded set of real numbers has a supremum, say $x$. Let, if possible,
$x < b$. Since $x \in [a, b]$, there is an open set, say $G_m$ that covers the
point $x$. If $G_m$ is of the form $(x - \delta, x + \delta)$ then the point
$x + \delta/2$ is also covered by $G_m$ and hence lies in $E$. But it is greater
than $\sup E$. So no such point exists and $x = b$.
\end{proof}

Its obvious generalisation is
\begin{thm}\label{s2t9}
A $k$-cell is compact.
\end{thm}
\begin{proof}
Let $a = (a_1, \ldots, a_k)$ and $b = (b_1, \ldots, b_k)$ be the corners of a
$k$-cell, $I_1$. The length of its diagonal is
\[
\delta = \left(\sum_{i=1}^k (b_i - a_i)^2\right)^{1/2} > 0,
\]
Let $\{G_\alpha\}$ be an open cover of the cell and it, if possible, there be no
finite subsover. Divide the cell into $2^k$ cells at the points $(a_i + b_i)/2$.
At least one of these cells, called $I_2$ will not have a finite subcover of
$\{G_\alpha\}$. We also note that $I_2 \subset I_1$. We divide $I_2$ is a 
similar fashion to get a cell with a diagonal $1/2^k$ times that of $I_2$ and
which does not have a finite subcover of $\{G_\alpha\}$. By theorem \ref{s2t7}, 
the sequences of these successively smaller, nested $k$-cells is non-empty. Let
$p \in \cap_{n \ge 1}I_n$. Then there exists a member, say $G_p$, in the open 
cover for which $B_\epsilon(p) \subset G_p$. 

If $p \in I_n$ and the diagonal of $I_n$ is $\delta_n$ then the maximum distance
between points of $I_n$ is $\delta_n$. If we choose $n$ such that 
\[
\delta_n = \frac{\delta}{2^{n}} < \epsilon
\]
the entire cell will be covered by $G_p$ alone, contradicting our choice of
choosing a sub-cell that can be covered only by an infinite subset of 
$\{G_\alpha\}$.

Therefore, our assumption that there is no finite subcover of $\{G_\alpha\}$
is false and the $k$-cell is compact.
\end{proof}

\begin{rem}
The proof of theorem \ref{s2t8} will be much simpler if conducted on lines of 
the previous proof.
\end{rem}

\begin{thm}\label{s2t10}
The following properties are equivalent for a set $E \subset \mathbb{R}^k$:
\begin{itemize}
\item $E$ is closed and bounded.
\item $E$ is compact.
\item Every infinite subset of $E$ has a limit point in $E$.
\end{itemize}
\end{thm}
\begin{proof}
If $E$ is closed and bounded then it can be enclosed in a $k$-cell. By theorem
\ref{s2t9} a $k$ cell is compact, making $E$ a closed subset of a compact set.
By theorem \ref{s2t3}, $E$ is compact.

If $E$ is compact then it is a compact subset of the metric space $\mathbb{R}^k$.
By theorem \ref{s2t3} it is closed. Therefore it contains all its limit points.

Now assume that every infinite subset of $E$ has a limit point in $E$ but $E$
is not bounded. Then there is a sequence of points $\{x_n \in E\}$ such that
for all $n \in \mathbb{N}$, $|x_n| > n$. Let $S$ be subset of $E$ with precisely
these points. Then $S$ has no limit point in $\mathbb{R}^k$ and therefore in 
$E$, contradicting our assumption. Therefore $E$ must be bounded.

Under the same assumptions as before, assume that $E$ is not closed. Then
there exists a limit point $x_0$ of $E$ that is not a member of $E$. Since 
$x_0$ is a limit point of $E$, for all $r > 0$, $B_r(x_0) \cap (E \setminus \{
x_0\}) \ne \varnothing$. If $x_0 \notin E$, then for all $r > 0$, $B_r(x_0) \cap
E \varnothing$. Define a set
\[
S = \{x_n \in E : x_n \in B_{1/n}(x_0) \cap E, x_n \ne x_m, \forall\; n, m \in
\mathbb{N}\}
\]
Clearly $x_0 \notin E$ is a limit point of $S \subset E$, a contradiction.
Therefore, $x_0$ must lie in $E$.
\end{proof}

Theorem \ref{s2t10} is a combination of
\begin{thm}[Heine-Borel]\label{s2t11}
For a set $E \subset \mathbb{R}^k$, $E$ is closed and bounded iff $E$ is compact.
\end{thm}
and
\begin{thm}[Weierstrass]\label{s2t12}
Every bounded infinite subet of $\mathbb{R}^k$ has a limit point in 
$\mathbb{R}^k$.
\end{thm}
\begin{proof}
Every bounded subset of $\mathbb{R}^k$ can be enclosed in a $k$-cell which is
closed.
\end{proof}

A theorem closely related to \ref{s2t12}
\begin{thm}[Bolzano-Weierstrass]\label{s2t13}
Every bounded sequence in $\mathbb{R}^k$ has a convergent subsequence.
\end{thm}
\begin{proof}
Let $S = \{x_n\}$ be a bounded sequence in $\mathbb{R}^k$. If $S$ is finite then
$x_n = p$ for infinitely many $n$. These elements form a subsequence of $S$ that
converges to $p$.

Now suppose $S$ is infinite. Then $S$ is a bounded subset of the metric space
$\mathbb{R}^k$. By Weirstrass theorem \ref{s2t12} it has a limit point, say $p$
in $\mathbb{R}^k$. Therefore, for all $r > 0$, $B_r(p) \cap (S - \{p\})$ is
non-empty. Create a sequence
\[
S_c = \{x_n \in S : x_n \in B_{1/n}(p) \cap (S - \{p\}), x_n \ne x_m, n, m \in 
\mathbb{N}\}.
\]
Clearly, $p$ is a limit point of $S_c$ and $S_c$ is a subsequence of $S$.
\end{proof}

The important theorems \ref{s2t11}, \ref{s2t12} and \ref{s2t13} are valid only
in $\mathbb{R}^k$ and not in general metric spaces. Their proofs depend on the
closeness and compactness of a $k$-cell. However we can retain a part of them.
\begin{defn}\label{s2d3}
A set $E$ is sequentially compact if every sequence in it has a convergent
subsequence.
\end{defn}
We now show that
\begin{thm}\label{s2t14}
A compact subset $E$ of a metric space $(X, d)$ is sequentially compact.
\end{thm}
\begin{proof}
Let $S$ be a sequence in a compact subset $E$ of a metric space $(X, d)$. Then
$S$ is an infinite subset of a compact set. By theorem \ref{s2t5}, $S$ has a 
limit point, say $p \in E$. Therefore, for all $n \ge 1$, $B_{1/n}(p) \cap (S
\setminus \{p\} \ne \varnothing$. Choose $x_n \in S$ such that for each $n \in
\mathbb{N}, x_n \ne x_m$ and $x_n \in B_{1/n}(p) \cap (S \setminus \{p\})$. 
The set $\{x_n\}$ form a convergent subsequence of $S$.
\end{proof}

In order to prove the converse, we need a few more definitions.
\begin{defn}\label{s2d4}
A set $E$ is totally bounded if for all $\epsilon > 0$, a finite set of open
balls of radius $\epsilon$ cover $E$.
\end{defn}

Consider the set of sequences $e_1 = (1, 0, 0, \ldots), e_2 = (0, 1, 0, \ldots),
e_3 = (0, 0, 1, \ldots), \ldots$. It is a metric space with the distance 
function 
\[
d(x, y) = \sqrt{\sum_{i \ge 1}(x_i - y-i)^2}.
\]
The set is bounded because $d(e_n, e_m) = \sqrt{2}$ for all $m, n$. But it is
not totally bounded because no finite set of balls of radii less than $\sqrt{2}$
will cover the set.

We now show that
\begin{thm}\label{s2t15}
Let $E$ be a subset of a metric space $(X, d)$ and let every sequence in $E$
have a convergent subsequence. Then $E$ is totally bounded.
\end{thm}
\begin{proof}
Assume that $E$ is not totally bounded. Then there exists $\epsilon > 0$
such that balls of radii less than or equal to $\epsilon$ do not cover the set.
Let $\{B_\epsilon(x_n)\}$ be the family of infinitely many balls that cover $E$.
Then $\{x_n\}$ is a sequence of points in $E$ without a convergent subsequence.
\end{proof}

It is easy to show that
\begin{thm}\label{s2t16}
If $E$ is a sequentially compact subset of a metric space then it is complete.
\end{thm}
\begin{proof}
Let $\{x_n\}$ be a Cauchy sequence in $E$. Then it has a convergent subsequence
with a limit point $p \in E$. Since $\{x_n\}$ is Cauchy $x_n \rightarrow p \in
E$, making $E$ complete.
\end{proof}

\begin{thm}\label{s2t17}
If $E$ is totally bounded and complete then $E$ is compact.
\end{thm}
\begin{proof}
Since $E$ is totally bounded, it can be covered with a finite number of balls 
of radius $1$. Let $\{G_\alpha\}$ be an open cover of $E$. Let, if possible, 
there be no finite sub-cover. Then there is at least one among those balls,
say $B_1$ that has no finite subcover in $\{G_\alpha\}$. Since $B_1$ is also
totally bounded, it can be covered with a finite number of balls of radius $1/2$.
Again, one of them, say $B_2$ will not have a finite subcover in $\{G_\alpha\}$.
Create a sequence of such balls $B_1 \supset B_2 \supset B_3 \supset \ldots$.
Build a sequence $S = \{x_n \in E: x_n \in B_n, x_n \ne x_m, \forall n, m \in
\mathbb{N}\}$. Since the radius of balls keeps shrinking, $S$ is a Cauchy
sequence. Since $E$ is complete, it converges to a point $p \in E$. Since $p$
is a limit point, it can be enclosed in a single open ball which can be covered
by a single member in $\{G_\alpha\}$. This contradicts the assumption that every
ball in the sequence needs an infinite subcover of $\{G_\alpha\}$. Therefore, 
we do not need an infinite subcover of $\{G_\alpha\}$ to cover $E$, making $E$
a compact set.
\end{proof}

We now have an extension of Heine-Borel theorem for arbitrary metric spaces.
\begin{thm}\label{s2t18}
If $E \subset (X, d)$ then $E$ is closed and totally bounded iff $E$ is compact.
\end{thm}
\begin{proof}
Assume that $E$ is compact. By theorem \ref{s2t14}, it is sequentially compact.
By theorem \ref{s2t16} it is closed and by theorem \ref{s2t15} it is totally 
bounded.

Now assume that $E$ is closed and totally bounded. By theorem \ref{s2t17} it is 
compact.
\end{proof}
\end{document}
